---
title: "Getting started"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
knitr::opts_chunk$set(eval = FALSE)
```


To use codriver you need access to a **large language model (LLM)**, either
through a cloud provider like OpenAI or Anthropic, a model running on your own
machine, or a proxy provided by your organisation. Codriver supports a wide
range of providers by building on the [ellmer package](https://ellmer.tidyverse.org).

This article explains how to connect codriver to your provider. It assumes you
already have access to one, meaning you have signed up, obtained an API key, or
been given credentials by your organisation. If you don't have access yet, see
[**Free API access**](https://vdwulp.github.io/codriver/articles/free-api-access.html)
for options that are free to start with.

------------------------------------------------------------------------

## Quick start

If you have an **API key set and know your provider**, this is all you need:

```{r}
# Install codriver
install.packages("codriver")

# OpenAI
codriver::codriver_configure("openai", model = "gpt-5.1")

# Anthropic Claude
codriver::codriver_configure("anthropic", model = "claude-sonnet-4-5")

# OpenAI-compatible endpoint
codriver::codriver_configure("openai_compatible",
  base_url = "https://llmproxy.example.com/v1",
  model    = "gpt-4.1"
)
```

Any provider and argument supported by [ellmer](https://ellmer.tidyverse.org)
works. Read on if you need help with provider naming, authentication, or
provider-specific notes.

------------------------------------------------------------------------

## Providers

Codriver does not communicate with LLM providers directly. Instead, it relies on
[**ellmer**](https://ellmer.tidyverse.org), a package that provides a unified
interface to a wide range of providers. Ellmer supports many popular services -
OpenAI, Anthropic Claude, Google Gemini, Mistral, Ollama, and more - and handles
the details of authentication and communication for each one.

When you call `codriver::codriver_configure()`, you tell codriver which provider
to use by passing its **name** — a short string like `"openai"` or
`"anthropic"`. Any argument accepted by the corresponding ellmer constructors
can be passed directly through `codriver_configure()`.

| Provider | `name` | Environment variable |
|---|---|---|
| OpenAI | `"openai"` | `OPENAI_API_KEY`<br>[see below](#openai) |
| OpenAI-compatible | `"openai_compatible"` | `OPENAI_API_KEY`<br>[see below](#openai-compatible-endpoint) |
| Azure OpenAI | `"azure_openai"` | `AZURE_OPENAI_API_KEY`<br>[see below](#azure-openai) |
| Anthropic | `"anthropic"` | `ANTHROPIC_API_KEY` |
| Google Gemini | `"google_gemini"` | `GOOGLE_API_KEY`<br>see [Free API access](https://vdwulp.github.io/codriver/articles/free-api-access.html#google-gemini) |
| AWS Bedrock | `"aws_bedrock"` | *(IAM credentials)* |
| Databricks | `"databricks"` | `DATABRICKS_TOKEN` |
| Groq | `"groq"` | `GROQ_API_KEY` |
| Ollama (local) | `"ollama"` | *(none required)*<br>see [Free API access](https://vdwulp.github.io/codriver/articles/free-api-access.html#ollama) |
| OpenRouter | `"openrouter"` | `OPENROUTER_API_KEY` |
| LM Studio (local) | `"lmstudio"` | *(none required)*<br>see [Free API access](https://vdwulp.github.io/codriver/articles/free-api-access.html#lm-studio) |
| Mistral | `"mistral"` | `MISTRAL_API_KEY`<br>see [Free API access](https://vdwulp.github.io/codriver/articles/free-api-access.html#mistral) |
| Perplexity | `"perplexity"` | `PERPLEXITY_API_KEY` |
| Snowflake | `"snowflake"` | `SNOWFLAKE_TOKEN` |
> _The table covers providers commonly used. Ellmer supports additional
providers. See the [ellmer homepage](https://ellmer.tidyverse.org/index.html#providers)
for the full and current list._

Most providers require a **model** to be specified, but ellmer will select a
default for most providers. If a default is used, ellmer will print it to the
console on every codriver call. Specifying the model explicitly in
`codriver_configure()` avoids this.

------------------------------------------------------------------------

## Authenticating

Before codriver can send requests to an LLM, it needs to prove to the provider
that you are authorised to use it. How this works depends on the provider.

The most common approach is an **API key**: a secret string that you obtain from
your provider's website and store on your machine. Ellmer looks for these keys
in **environment variables** — named slots in your R session that hold
configuration values. Each provider expects its key in a specific variable.

The table above lists the environment variable names of commonly used providers.
If your provider is not in the table or uses a different authentication
mechanism, look at the help page for the corresponding ellmer constructor. For
example, running `?ellmer::chat_openai` in R will tell you that OpenAI expects
`OPENAI_API_KEY` and `?ellmer::chat_aws_bedrock` tells you about their specific
authentication mechanism.

The recommended way to set an environment variable permanently is to add it to
your `.Renviron` file, which R reads automatically at startup. The OpenAI
example below shows how this is done.

------------------------------------------------------------------------

## OpenAI

OpenAI is the provider codriver has been most thoroughly tested with. GPT-4.1
and GPT-5.1 both produce clean, reliable completions in the RStudio workflow
codriver is designed for.

To get started, go to [platform.openai.com](https://platform.openai.com), sign
in, and create an API key under **API keys**.

Then store it in your `.Renviron`:

```{r}
file.edit("~/.Renviron")
```

Add:

```         
OPENAI_API_KEY=your_key_here
```

Save the file, *restart RStudio* and then configure codriver:

```{r}
codriver::codriver_configure("openai", model = "gpt-5.1")
```

That is all that is needed. You're now ready to
[start using codriver](start-using-codriver.html) in your RStudio workflow.

------------------------------------------------------------------------

## OpenAI-compatible endpoint

Many organisations run their own LLM infrastructure that speaks the OpenAI API
format. This includes local model servers, university research clusters, and
company proxies. For these, use `"openai_compatible"` as the provider name.

Unlike named providers, an OpenAI-compatible endpoint has no defaults. You will
need to know the **base URL** and **model name** from your organisation's
documentation or administrator.

By default, `"openai_compatible"` reads the API key from the environment
variable `OPENAI_API_KEY`. If you want to keep it separate from an actual
OpenAI key, save a different key under a custom environment variable name and
pass it explicitly:

```{r}
codriver::codriver_configure(
  "openai_compatible",
  base_url    = "https://llmproxy.example.com/v1",
  model       = "gpt-5.1",
  credentials = function() Sys.getenv("CUSTOM_API_KEY")
)
```

> _Note: older ellmer documentation may show an `api_key` argument - this is
deprecated in favour of `credentials` in combination with an environment
variable, which avoids hardcoding keys in your code._

------------------------------------------------------------------------

## Microsoft Azure OpenAI

Another service regularly used by organisations is `"azure_openai"`. For this
provider, you always need the **endpoint** of your Azure resource. In addition,
authentication is handled in one of three ways. Your IT team can tell you which
authentication method applies and what values to use.

##### Option 1: API key

Set the endpoint and API key in your `.Renviron`:
```r
file.edit("~/.Renviron")
```

Add:
```
AZURE_OPENAI_ENDPOINT=https://your-resource.openai.azure.com
AZURE_OPENAI_API_KEY=*your_key_here*
```

Save the file, *restart RStudio* and then configure codriver (see below).

##### Option 2: Service principal

Set the endpoint and service principal values in your `.Renviron`:
```r
file.edit("~/.Renviron")
```

Add:
```
AZURE_OPENAI_ENDPOINT=https://your-resource.openai.azure.com
AZURE_TENANT_ID=*your_tenant_id*
AZURE_CLIENT_ID=*your_client_id*
AZURE_CLIENT_SECRET=*your_client_secret*
```

Save the file, *restart RStudio* and then configure codriver (see below).

##### Option 3: Entra ID

Set the endpoint in your `.Renviron`:
```r
file.edit("~/.Renviron")
```

Add:
```
AZURE_OPENAI_ENDPOINT=https://your-resource.openai.azure.com
```

Save the file, *restart RStudio* and then configure codriver (see below). When
none of the other authentication options are used, an *Entra ID* browser login
will open when configuring codriver.

##### Configure codriver

To configure codriver after setting authentication values, use:
```r
codriver::codriver_configure("azure_openai", model = "your-deployment-name")
```

------------------------------------------------------------------------

## Provider notes

Some notes on experience with providers:

- **OpenAI** is the provider codriver has been most thoroughly tested with.
  GPT-4.1 and GPT-5.1 both produce clean, reliable completions in the RStudio
  workflow codriver is designed for.
- **Anthropic Claude** models have also been tested quite extensively.
  Especially Claude Sonnet 4.5 has been confirmed to work well with codriver.
- **Google Gemini** produces good results too. When using _free tier_ be aware
  that some models are _very slow_ and only allow for a few requests per day.
  Gemini 3.1 Flash Lite and Gemini 3.5 Flash Lite have been proven to be quick
  and working well with codriver. See [Free API access](https://vdwulp.github.io/codriver/articles/free-api-access.html#google-gemini)
  for more details.
  
  > The Gemini Flash 2.5 (non-lite) series models use an internal reasoning
  process that can occasionally cause thought text to appear in completions.
  If you encounter this, try passing `api_args` to suppress thought output in
  the response:
  
  ```{r}
  codriver::codriver_configure(
    "google_gemini",
    model    = "gemini-2.5-flash",
    api_args = list(
      generationConfig = list(
        thinkingConfig = list(includeThoughts = FALSE)
      )
    )
  )
  ```

------------------------------------------------------------------------

**Next: [Using codriver](start-using-codriver.html)**
